Supplementary Information from:\\ \textit{Reconstructing the last known movements of one of Nature's giants}

### 2017-11-07 manuscript/revision/blue-whale-supplementary-v2.tex
Typo correction for referee 1

\subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
The baleen plate yielded 78 discrete samples of baleen. 
Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower. 
Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
$\delta^{13}$C values also show repeated fluctuations in the first (oldest) 60cm of the baleen plate (behavioural phase one), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two), with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.

% figure s1
\begin{figure}[!htbp]
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
  \caption{Fourier analysis of $\delta^{15}$N (A,B) and $\delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate. Inset values are the two highest frequency spectral densities (in cm). Note the baleen plate is 80cm long. Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two} 
  \label{figs1}
\end{figure}

Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioral phase one (oldest 60cm of the baleen plate; Fig. \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioral phase two (the final 20cm of the baleen plate).

% figure s2
\begin{figure}[!htbp]
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm} % check legend
  \label{figs2}
\end{figure}

@@

Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Fig. \ref{figs3}B). 
Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}A). 
The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the initial phase of the baleen plate are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months. 
Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d15N-map.png} 
    \end{subfigure}
    \caption{Spatial model (isoscape) of model-simulated biomass weighted annual average (A) $\delta^{13}$C and (B) $\delta^{15}$N values from POM. $\delta^{13}$C values are from Magozzi et al. 2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and Somes (pers. comm.)} 
    \label{figs3}
\end{figure}

@@

\subsection*{ISOTOPE MODEL SIMULATIONS}
We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions of that would be encountered by whales foraging within broad geographic regions (Norwegian Sea, west UK/Irish shelf, Canaries/west Azores, Mid Atlantic ridge; Fig. \ref{figs4}). 
Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Fig. \ref{figs4}). 
In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Fig. \ref{figs4}), whereas in subtropical regions exemplified by the Canaries and western Azores/Mid Atlantic ridge areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig. \ref{figs4}).

  \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to Norway, Ireland, Canaries and the Mid Atlantic ridge as labelled over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
      \label{figs4}
  \end{figure}

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes are similar to the measured $\delta^{13}$C profile during the four years of behavioral phase one. 
Simulated profiles display the same asymmetric isotopic cycles associated with development of the spring bloom, with annual $\delta^{13}$C maxima in June (Fig. \ref{figs4}, Fig. 1). 
Simulated temporal isotopic profiles for northern resident whales, do not display clear isotopic minima preceding the onset of the bloom. 
Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig. \ref{figs4}, Fig. 1).

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
Behavioural phase two is characterized by a dramatic shift to maximum $\delta^{13}$C values in spring 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values. 
$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890. 
The final sample representing feeding shortly before death in Irish waters implies a return to isotopic conditions more consistent with behavioural phase one. 
The relatively positive and constant carbon isotope composition of behavioural phase two is consistent with residency (and continued feeding) in more southerly latitudes, particularly in the relatively isotopically positive waters southwest of the Azores.

Accordingly we modelled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the late spring of 1889, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 1). 
Consistently low $\delta^{15}$N values in 1890 are also consistent with residency in relatively southern waters.

@@

\subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by SST and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig. \ref{figs5}).

The agent-based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
At each daily step, the probability of movement, movement directions and distance traveled are sampled from probability distributions to allow individual variation.

@@

\centering
\begin{table}
  \begin{tabular}{|p{5cm}|c|c|c|c|} 
    \hline
    & \multicolumn{4}{|c|}{\textbf{Behavioural state}} \\
    \hline
    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south} & \textbf{Nursing}\\
    \hline
    Month - behavioural phase one (1460 days) & May - Oct & Feb - April & Nov - Jan & NA\\
    \hline
    Month - pregnancy (365 days) & June - Oct & April - May & Nov - March & NA\\
    \hline
    Month - nursing (210 days) & NA & NA & NA & June - Dec\\
    \hline
    Minimum sea surface temperature ($^{\circ}$C) & 3 & 3 & 3 & 18\\
    \hline
    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25 & 25\\
    \hline
    Maximum daily distance traveled (km) & $N(\mu=30, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=20, \sigma=5)$\\
    \hline
    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA & NA\\
    \hline
  \end{tabular}
  \caption{Model parameters for agent-based model of whale movement. See Fig. \ref{figs5} for details.}
  \label{tables1}
\end{table}

>>>>

\subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
The baleen plate yielded 78 discrete samples of baleen. 
Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower. 
Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
$\delta^{13}$C values also show repeated fluctuations in the first (oldest) 60cm of the baleen plate (behavioural phase one), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two, with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.

% figure s1
\begin{figure}[!htbp]
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
  \caption{Fourier analysis of $\delta^{15}$N (A,B) and $\delta^{13}$C (C,D) values of the full (A,C) and behavioural phase one only (B,D) components of the baleen plate. Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e. 80cm), and at the sharp increase of $\delta^{13}$C seen at 60cm. Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two.} 
  \label{figs1}
\end{figure}

Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate; Fig. \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase two (the final 20cm of the baleen plate).

% figure s2
\begin{figure}[!htbp]
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm.} % check legend
  \label{figs2}
\end{figure}

@@

Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Fig. \ref{figs3}). 
Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the initial phase of the baleen plate are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months. 
Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d15N-map.png} 
    \end{subfigure}
    \caption{Spatial model (isoscape) of model-simulated biomass weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from POM. $\delta^{13}$C values are from Magozzi et al. 2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and C.J. Somes (pers. comm.).} 
    \label{figs3}
\end{figure}

@@

\subsection*{ISOTOPE MODEL SIMULATIONS}
We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge; Fig. \ref{figs4}). 
Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Fig. \ref{figs4}). 
In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Fig. \ref{figs4}), whereas in subtropical regions exemplified by the Canaries and western Azores/Mid Atlantic ridge areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig. \ref{figs4}).

  \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to west UK/Irish shelf (Ireland), Norwegian/Barents Sea (Norway), Canaries/west Azores (Canaries), Mid Atlantic ridge (Mid-Atlantic), over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
      \label{figs4}
  \end{figure}

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes are similar to the measured $\delta^{13}$C profile during the four years of behavioural phase one. 
Simulated profiles display the same asymmetric isotopic cycles associated with development of the spring bloom, with annual $\delta^{13}$C maxima in June (Fig. \ref{figs4}, Fig. 3). 
Simulated temporal isotopic profiles for northern resident whales, do not display clear isotopic minima preceding the onset of the bloom (Fig. \ref{figs4}). 
Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig. \ref{figs4}, Fig. 1).

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
Behavioural phase two is characterised by a dramatic shift to maximum $\delta^{13}$C values in spring 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values. 
$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890 (Fig. 1). 
The final sample representing feeding shortly before death in Irish waters implies a return to isotopic conditions more consistent with behavioural phase one. 
The relatively positive and constant carbon isotope composition of behavioural phase two is consistent with residency (and continued feeding) in more southerly latitudes, particularly in the relatively isotopically positive waters southwest of the Azores.

Accordingly we modeled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the late spring of 1889, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 3). 
Consistently low $\delta^{15}$N values in 1890 are also consistent with residency in relatively southern waters.

@@

\subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig. \ref{figs5}).

The agent-based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
At each daily step, the probability of movement, movement direction and distance traveled are sampled from probability distributions to allow individual variation.

@@

\centering
\begin{table}
  \begin{tabular}{|p{5cm}|c|c|c|c|} 
    \hline
    & \multicolumn{4}{|c|}{\textbf{Behavioural state}} \\
    \hline
    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south} & \textbf{Nursing}\\
    \hline
    Month - behavioural phase one (1460 days) & May - Oct & Feb - April & Nov - Jan & NA\\
    \hline
    Month - pregnancy (365 days) & June - Oct & April - May & Nov - March & NA\\
    \hline
    Month - nursing (210 days) & NA & NA & NA & June - Dec\\
    \hline
    Minimum sea surface temperature ($^{\circ}$C) & 3 & 3 & 3 & 18\\
    \hline
    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25 & 25\\
    \hline
    Maximum daily distance traveled (km) & $N(\mu=30, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=20, \sigma=5)$\\
    \hline
    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA & NA\\
    \hline
  \end{tabular}
  \caption{Model parameters for agent-based model of whale movement. See Fig. \ref{figs5} and text for more details.}
  \label{tables1}
\end{table}
